Staged Abstract Interpreters

### 2019-09-05 paper/10_intro.tex
intro

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes the program by traversing the abstract syntax tree (AST). The traversal incurs
interpretive overhead, such as the pattern matching on the ASTs and recursive
calls on subexpressions. Such overhead can be negligible if the abstract
interpreter only runs on the program a single time, but can accumulate
significantly if it runs repeatedly (for example, on libraries).  Second, it is
common that abstract interpreters written in pure languages extensively use
functional encodings of side-effects to implement the semantics of abstract interpretation.
For example, the list monad can be used to model nondeterminism from collecting
interpreters.
%For example, a collecting interpreter that returns all possible runtime values is
%nondeterministic, which can be modeled by a list monad.
Although such pure approaches have their own merits, compared with imperative,
stateful implementations, they are significantly slower.  In order to
reconcile these approaches, one must use a programming technique which allows
for high-level abstractions during implementation, but bypasses the
incurred overhead at runtime.

Roughly at the same time when abstract interpretation was proposed in
1970s, \citet{futamura1971partial} revealed a close
connection between interpreters and compilers through a hierarchy of
specializations, namely, Futamura Projections. The first Futamura
projection shows that specializing an interpreter to the input program
removes interpretative overhead and yields an equivalent executable
that runs faster. The Futamura projections have been successfully applied to
other similar performance problems, examples including
query compilers\cite{DBLP:conf/sigmod/TahboubER18}, PyPy \footnote{PyPy:
\url{http://pypy.org/}}, Truffle \cite{Marr:2015:TVP:2814270.2814275}, and etc.
The challenge here: can we apply the same idea to a general abstract interpreter
to make it run faster? To achieve this, we need to specialize a nonstandard
semantics that extensively uses nondeterminism, operates on abstract domains,
and computes fixed-points iteratively over a cache. The solution is not simply obvious.

@@

The first Futamura projection specifically shows that specializing an
interpreter w.r.t. the input program yields an equivalent executable. For
instance, @eval : Expr -> Input -> Value@ is an interpreter for some language.
Given a program $e_0$ of type @Expr@, by applying the specialization, we can
obtain a specialized interpreter $\texttt{eval}_{\texttt{e0}}$ : @Input -> Value@,
which is the so-called \textit{equivalent executable}.

@@

The first proposed approach to realize Futamura projections is partial
evaluation \cite{DBLP:books/daglib/0072559}, which relies on an automatic
binding-time analysis (\textit{static} or \textit{dynamic}).  The underlying
representation can be viewed as a \textit{two-level semantics}
\cite{NIELSON1989117, NIELSON198859} that distinguishes compile-time
computation and run-time computation.
However, it is hard to precisely analyze binding-times given an arbitrary
program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} allows users to explicitly program over
a two-level syntax \cite{Nielson:1992:TFL:130665}, by manually
annotating the binding-times of variables in the program. Then, the MSP system
will check whether these annotations are consistent, and specialize the program
using that information. The staging annotations can be either syntactic (e.g.,
quote and quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight
Modular Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

To apply Futamura projections on abstract interpreters, we consider
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s big-step monadic abstract
interpreter as the unstaged baseline.  Similar to the two-level semantics
\cite{NIELSON1989117}, monads provide a good abstraction to hide the detail of
abstract interpretation semantics. However, the monads do not have a clear
stage distinction.
We further introduce binding-times and make the monads binding-time polymorphic.
Our proposed framework adopts the type-based multi-stage programming from the
Lightweight Modular Staging framework, and implements the Futamura projection
of a big-step abstract interpreter for a small, stateful, higher-order
language. By deriving staged monads that can be used to do code generation, we
obtain a staged abstract interpreter that can be used for specialization. After
staging, the generated code is specialized to the input program, and all the
monadic operations are inlined and compiled down to low-level Scala code.

@@

Although the four artifacts may look dissimilar at first glance, they in
fact are all firmly rooted in the concrete semantics of the language.  This
observation provides a way to abstract over the interpreter and achieve the
flexibility of reinterpreting the shared interpreter.  The generic interpreter
returns a value of monadic type, which can be varied by different semantics.
The domain of the interpreter and the effects such as read, state, and
nondeterminism can all be wrapped into this monadic type.  The binding-time
abstraction is represented by a higher-kinded type, and controls whether the
interpreter produces values directly or generates code. It is worth mentioning
that the binding-time type is also injected into the monadic type, so that we
will distinguish normal monads and staged monads.

\paragraph{Applications and Evaluation}
We evaluate the idea of staging an abstract interpreter through
case studies and an empirical evaluation on performance.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, an implementation technique for
control-flow analyses. We show that by utilizing type-based stage
annotations, we can achieve the same optimizations. Meanwhile,
the analyzer does not need to change, except the addition of stage annotations,
thereby requiring significantly less engineering effort.
2) We extend the basic staged abstract interpreter to different flow
analyses, including a store-widened analysis, a context-sensitive
analysis, and abstract garbage collection.
3) We show that staging an abstract interpreter enables modular
compilation of an analysis to programs. Here we borrow the concept of
modular analysis, and show that the compiled analysis is reusable.
Therefore, the approach provides a modular way to create optimized
analysis code by mechanized reuse of a whole-program analyzer.
4) We empirically evaluate the performance improvements gained by staging,
showing an order of magnitude speedup on flow analyses.

>>>>

However, besides the intrinsic complexity of static analysis, there are
additional abstraction penalties from these high-level implementation
approaches. First, similar to concrete interpreters, an abstract interpreter
analyzes the program by traversing the AST. The traversal incurs
interpretive overhead, such as the pattern matching on the ASTs and recursive
calls on subexpressions. Such overhead can be negligible if the abstract
interpreter only runs on the program a single time, but can accumulate
significantly if it runs repeatedly (for example, on libraries).  Second, it is
common that abstract interpreters written in pure languages extensively use
functional encodings of side-effects to implement the semantics of abstract interpretation.
For example, the list monad can be used to model nondeterminism from collecting
interpreters.
%For example, a collecting interpreter that returns all possible runtime values is
%nondeterministic, which can be modeled by a list monad.
Although such pure approaches have their own merits, compared with imperative,
stateful implementations, they are significantly slower.  In order to
reconcile these approaches, one must use a programming technique which allows
for high-level abstractions during implementation, but bypasses the
incurred overhead at runtime.

Roughly at the same time when abstract interpretation was proposed in
1970s, \citet{futamura1971partial} revealed a close
connection between interpreters and compilers through a hierarchy of
specializations, namely, Futamura Projections. The first Futamura
projection shows that specializing an interpreter to the input program
removes interpretative overhead and yields an equivalent executable
that runs faster. The Futamura projections have been successfully applied to
other similar performance problems, examples including
query compilers\cite{DBLP:conf/sigmod/TahboubER18}, PyPy \footnote{PyPy:
\url{http://pypy.org/}}, Truffle \cite{Marr:2015:TVP:2814270.2814275}, and etc.
The challenge here: can we apply the same idea to a general abstract interpreter
to make it run faster? The solution is not obvious, as we need to specialize
a nonstandard semantics that extensively uses nondeterminism, operates on
abstract domains, and computes fixed-points iteratively over a cache.

@@

The first Futamura projection specifically shows that specializing an
interpreter w.r.t. the input program yields an equivalent executable. For
instance, @eval : Expr -> Input -> Value@ is an interpreter for some language.
Given a program $e_0$@ : Expr@, by applying the specialization, we can
obtain a specialized interpreter $\texttt{eval}_{\texttt{e0}}$ : @Input -> Value@,
which is the so-called \textit{equivalent executable}.

@@

The first proposed approach to realize Futamura projections is partial
evaluation \cite{DBLP:books/daglib/0072559}, which relies on an automatic
binding-time analysis (\textit{static} or \textit{dynamic}).  The underlying
representation can be viewed as a \textit{two-level semantics}
\cite{NIELSON1989117, NIELSON198859} that distinguishes compile-time
computation and run-time computation.
However, it is hard to precisely analyze binding-times given an arbitrary
program. As an alternative approach, multi-stage programming (MSP)
\cite{taha1999multi, DBLP:conf/pepm/TahaS97} allows users to explicitly program over
a two-level syntax \cite{Nielson:1992:TFL:130665}, by manually
annotating the binding-times of variables. Then, the MSP system
will check whether these annotations are consistent, and specialize the program
using that information. The staging annotations can be either syntactic (e.g.,
quote and quasi-quote in MetaML/MetaOCaml) or type-based (e.g., the Lightweight
Modular Staging framework \cite{DBLP:conf/gpce/RompfO10} in Scala).

To apply Futamura projections on abstract interpreters, we consider
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s big-step monadic abstract
interpreter as the unstaged baseline.  Similar to the two-level semantics
\cite{NIELSON1989117}, monads provide a good abstraction to hide the detail of
abstract interpretation semantics. However, the monads do not have a clear
stage distinction.
We further introduce binding-times and make the monads binding-time polymorphic.
Our proposed framework adopts the type-based multi-stage programming from the
Lightweight Modular Staging framework, and implements the Futamura projection
of a abstract interpreter for a stateful higher-order
language. By deriving staged monads that can be used to do code generation, we
obtain a staged abstract interpreter that can be used for specialization. After
staging, the generated code is specialized to the input program, and all the
monadic operations are inlined and compiled down to low-level Scala code.

@@

Although the four artifacts may look dissimilar at first glance, they in
fact are all firmly rooted in the concrete semantics of the language.  This
observation provides a way to abstract over the interpreter and achieve the
flexibility of reinterpreting the shared interpreter.  The generic interpreter
returns a value of monadic type, which can be varied by different semantics.
The value domains of the interpreter and the effects such as state and nondeterminism
can all be wrapped into this monadic type.  The binding-time abstraction is
represented by a higher-kinded type, and injected into the monadic type.  The
binding-times will control whether the interpreter produces values directly or
generates code.

%It is worth mentioning
%that the binding-time type is also injected into the monadic type, so that we
%will distinguish normal monads and staged monads.

\paragraph{Applications and Evaluation}
We evaluate the idea of staging an abstract interpreter through
case studies and an empirical evaluation on performance.
1) We compare our approach with abstract compilation
\cite{Boucher:1996:ACN:647473.727587}, an implementation technique for
control-flow analyses. We show that by utilizing type-based stage
annotations, we can achieve the same optimizations. Meanwhile,
the analyzer does not need to change, except the addition of stage annotations,
thereby requiring significantly less engineering effort.
2) We extend the basic staged abstract interpreter to different flow
analyses, including a store-widened analysis, a context-sensitive
analysis, and abstract garbage collection.
3) We show that staging an abstract interpreter enables modular
compilation of an analysis to programs. Here we borrow the concept of
modular analysis, and show that the compiled analysis is reusable.
Therefore, the approach provides a modular way to create optimized
analysis code by mechanized reuse of a whole-program analyzer.
4) We empirically evaluate the performance improvements gained by staging,
showing an order of magnitude speedup.
